DPhil Thesis Proposal: Semantics of Probabilistic Functional Languages

### 2021-03-30 thesis proposal/thesis proposal.tex
Add references

\documentclass[titlepage]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{stmaryrd}
\newcommand{\tY}{\mathsf{Y}}
\newcommand{\tsample}{\mathsf{sample}}
\newcommand{\tscore}{\mathsf{score}}
\newcommand{\denotation}[1]{\llbracket #1 \rrbracket}

@@

@@

Alternatively, the distribution of results can be specified more directly as a distribution. This omits any information on joint distributions from the semantics, but this itself can be in some ways an advantage, as a subterm can be duplicated without being correlated to itself. It would be natural to attempt to use the category of measurable spaces, however, that turns out to not be cartesian closed. This motivates the introduction of quasi-Borel spaces as an alternative way in which a distribution can be defined, which has a more restrictive set of morphisms and does not have this issue, but is nevertheless still able to represent all the functions definable as programs.

@@

\subsection{Operational Semantics} \label{operational semantics}
In operational semantics, the meaning of the program is defined by repeatedly transforming a term into other terms, until it reaches a value, a term which is considered finished, which is the output. A small-step trace-based operational semantics of the example language can be defined as follows.

@@

\subsection{Denotational Semantics}
The general format of denotational semantics is a function $\denotation M_\gamma$, from a term $M$ to its value, given the values of all the free variables in the context $\gamma$. The type of output depends on the type of the term, given by $\denotation A$ which is defined recursively. For a non-probabilistic language, it would be something like $\denotation{\mathsf T} = T$, $\denotation{A \to B} = \denotation B^{\denotation A}$. If $M$ satisfies the typing judgement $\Gamma \vdash M : A$, and $\gamma \in \denotation \Gamma$ (i.e.~$\gamma$ is a function from the variables in $\Gamma$ to values such that if $x:B$ is in $\Gamma$, $\gamma (x) \in \denotation B$), then $\denotation M_\gamma \in \denotation A$.

For the example language but without the probabilistic constructs (i.e.~just simply-typed lambda calculus with constants), a denotational semantics could be defined as
\begin{align*}
\denotation t_\gamma & = t\,(\text{for }t \in T) \\
\denotation x_\gamma & = \gamma(x) \\
\denotation{M N}_\gamma & = \denotation M_\gamma(\denotation N_\gamma) \\
\denotation{\lambda x. M }_\gamma(X) & = \denotation M_{\gamma ; x \mapsto X}.
\end{align*}
In order to define a denotational semantics for a probabilistic language, the result would have to be not just a value, but some sort of distribution of values. The rule $\denotation{A \to B} = \denotation B^{\denotation A}$ also has to be changed to take into account that a function may produce outputs that differ randomly each time it is called, so something like the set of kernels from $\denotation A$ to $\denotation B$ (a trace-based denotational semantics would also probably be possible, but much more complicated, but as far as I am aware, such a thing has never been defined). This would require a measure structure on $\denotation A$ and $\denotation B$ to be defined, but in some cases it is not possible to define a measurable structure on the set of kernels between two measurable spaces such that the application map taking a kernel $A \rightsquigarrow B$ and an element of A is itself measurable, therefore the denotations of higher-order types could be undefined. This motivates the introduction of quasi-Borel spaces.

@@

Having a denotational semantics available makes certain sorts of proofs more convenient. For example, if there's a program rearrangement that can be expressed as replacing $A[M,N]$ with $B[M,N]$ (for arbitrary $M, N$), then the proof of equivalence can be quite direct, showing that the way $\denotation M$ and $\denotation N$ are combined is equivalent (possibly subject to some conditions on certain variables not being free in $M$ and $N$, which implies that $\denotation M_\gamma$ and $\denotation N_\gamma$ are independent of the values of those variables in $\gamma$), whereas with operational semantics, it is much more complicated to keep track of the way the arbitrary subterms are duplicated, transformed and combined as execution progresses. Similarly, the notion of contextual equivalence is not necessary in denotational semantics (being trivially equivalent to equality of values).

Another useful property of denotation semantics is that termination properties are much more directly visible. If the type system is strong enough to imply termination, this manifests directly as every term being given a value. If non-terminating terms are possible, it will be necessary to introduce, as part of the semantics, special cases to deal with that, otherwise the definition will simply not be consistent.

@@

There are a few variations on how the traces work. In some versions (as is the case in Section \ref{operational semantics}), each run of the program is assigned a trace of finite length, containing exactly those samples actually used in that run. If the set of results for an individual sampling operation is $S$ (in this case it's the set of 2 elements, but in a language with continuous distributions, it could be $\mathbb R$, or an interval, or something else), the set of traces is $\bigcup_{n \in \mathbb N} S^n$, and the measure defined on the set of traces is not actually a probability measure, as each set $S^n$ has measure 1. This still acts somewhat like a probability measure though, as traces of different lengths are mutually exclusive. If a trace $a$ is a prefix of a trace $b$ (with $b = a + c$), then it is not possible that $(M, a, 1) \to^* (V, [], s)$ and also $(M, b, 1) \to^* (V', [], s')$, because the evaluation with any given trace is deterministic, and $(M, b, 1) \to^* (V, c, s)$, at which point it could not proceed any further, having already reached a value. For any term, the set of traces for which it terminates therefore has measure at most 1.

Alternatively, the trace could contain infinitely many values, with any finite execution of a program leaving some of the samples unused. This is in some ways a little simpler, because the set of traces can simply have a probability distribution, and the trace does not need to depend in any way on the program (by having its length matched). An additional variant of this style of trace is to split the trace in three parts, all infinitely long, for the application case of the big-step semantics, rather than having the each step (evaluating the function, evaluating the argument, then evaluating the result of the $\beta$ reduction) use however many samples it needs from the start of the trace in turn.

@@

@@

The second extension is antitone ranking functions. This idea is taken from \cite{some citation needed}, just adapted to the new functional context. A function $\epsilon : \mathbb R_{\geq 0} \to \mathbb R_{> 0}$, which is required to be monotoically decreasing, is added to the definition, and rather than the ranking function having to decrease by 1 at every $\tY$-reduction step, it only has to decrease by $\epsilon$ of the current value of the ranking function. This allows the termination of programs which terminate much more slowly (such as the unbiased random walk) to be proven, whereas the basic result was only applicable to terms which terminated with a finite expected number of $\tY$-reduction steps.

@@

\subsection{Completeness of Antitone Rankability}
A term is rankable iff every term reachable from it terminates with finitely many $\tY$-reduction steps in expectation. This essentially puts a limit on how slowly a term can terminate while still being rankable, so any term which is not $\tY$-positive almost surely terminating cannot be proven AST by assigning it a ranking function. With antitone ranking functions, there is no analogous restriction: there are arbitrarily slowly terminating terms (i.e. for any probability distribution on $\mathbb N$, there is a term whose distribution of number of $\tY$-reduction steps before termination is the given distribution) which are antitone rankable. It therefore seems rather likely that any term such that every term reachable from is is AST is antitone rankable. I intend to try to prove this result, even though it does not relate very closely to my overall thesis topic, because it would make my existing work more complete.
\end{document}

>>>>

\documentclass[titlepage]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{natbib}
\setcitestyle{numbers, open={[}, close={]}}
\bibliographystyle{plainnat}
\usepackage{stmaryrd}
\newcommand{\tY}{\mathsf{Y}}
\newcommand{\tsample}{\mathsf{sample}}
\newcommand{\tscore}{\mathsf{score}}
\newcommand{\denotation}[1]{\llbracket #1 \rrbracket}

@@

\usepackage{hyperref}

@@

Alternatively, the distribution of results can be specified more directly as a distribution. This omits any information on joint distributions from the semantics, but this itself can be in some ways an advantage, as a subterm can be duplicated without being correlated to itself. To define a denotational semantics for a higher-order probabilistic language, it would be natural to attempt to use the category of measurable spaces, however, that turns out to not be cartesian closed. This motivates the introduction of quasi-Borel spaces \cite{8005137} as an alternative way in which a distribution can be defined, which has a more restrictive set of morphisms and does not have this issue, but is nevertheless still able to represent all the functions definable as programs.

@@

\subsection{Operational Semantics} \label{operational semantics}
In operational semantics, the meaning of the program is defined by repeatedly transforming a term into other terms, until it reaches a value, a term which is considered finished, which is the output. Several variants of operational semantics can be seen, for example, in \cite{borgstrom2016lambda}. A small-step trace-based operational semantics of the example language can be defined as follows.

@@

\subsection{Denotational Semantics}
The general format of denotational semantics is a function $\denotation M_\gamma$, from a term $M$ to its value, given the values of all the free variables in the context $\gamma$. The type of output depends on the type of the term, given by $\denotation A$ which is defined recursively. For a non-probabilistic language, it would be something like $\denotation{\mathsf T} = T$, $\denotation{A \to B} = \denotation B^{\denotation A}$.\cite{barendregt2013lambda} If $M$ satisfies the typing judgement $\Gamma \vdash M : A$, and $\gamma \in \denotation \Gamma$ (i.e.~$\gamma$ is a function from the variables in $\Gamma$ to values such that if $x:B$ is in $\Gamma$, $\gamma (x) \in \denotation B$), then $\denotation M_\gamma \in \denotation A$.

A key property that makes it denotational semantics (the result of the operational semantics given above could be reformatted into something like this, for example) is that the meaning of each term is defined in terms of the meanings of its subterms. \cite[\S 3.1]{MOSSES1990575}

For the example language but without the probabilistic constructs (i.e.~just simply-typed lambda calculus with constants), a denotational semantics could be defined as
\begin{align*}
\denotation t_\gamma & = t\,(\text{for }t \in T) \\
\denotation x_\gamma & = \gamma(x) \\
\denotation{M N}_\gamma & = \denotation M_\gamma(\denotation N_\gamma) \\
\denotation{\lambda x. M }_\gamma(X) & = \denotation M_{\gamma ; x \mapsto X}.
\end{align*}
In order to define a denotational semantics for a probabilistic language, the result would have to be not just a value, but some sort of distribution of values. The rule $\denotation{A \to B} = \denotation B^{\denotation A}$ also has to be changed to take into account that a function may produce outputs that differ randomly each time it is called, so something like the set of kernels from $\denotation A$ to $\denotation B$ (a trace-based denotational semantics would also probably be possible, but much more complicated, but as far as I am aware, such a thing has never been defined). This would require a measure structure on $\denotation A$ and $\denotation B$ to be defined, but in some cases it is not possible to define a measurable structure on the set of kernels between two measurable spaces such that the application map taking a kernel $A \rightsquigarrow B$ and an element of A is itself measurable, therefore the denotations of higher-order types could be undefined. This motivates the introduction of quasi-Borel spaces.\cite{8005137}

@@

Having a denotational semantics available makes certain sorts of proofs more convenient. Suppose there's a program rearrangement that can be expressed as replacing $A[M,N]$ with $B[M,N]$ (for arbitrary $M, N$), then the proof of equivalence can be quite direct, showing that the ways $\denotation M$ and $\denotation N$ are combined by $A$ and $B$ are equivalent (possibly subject to some conditions on certain variables not being free in $M$ and $N$, which implies that $\denotation M_\gamma$ and $\denotation N_\gamma$ are independent of the values of those variables in $\gamma$), whereas with operational semantics, it is much more complicated to keep track of the way the arbitrary subterms are duplicated, transformed and combined as execution progresses. For example, in \cite{staton2017commutative}, the equivalence of $\mathsf{let}\ x = A\ \mathsf{in\ let}\ y = B\ \mathsf{in}\ C$ and $\mathsf{let}\ y=B\ \mathsf{in\ let}\ x=A\ \mathsf{in}\ C$ was proven using a denotational semantics (subject to $x$ and $y$ not being free in $A$ and $B$). Similarly, the notion of contextual equivalence is not generally necessary in denotational semantics, being trivially equivalent to equality of values (though there are some cases where it could still apply, if some class of terms do not count as valid complete programs).

Another useful property of denotation semantics is that termination properties, to the extent that they exist, are much more directly visible. If the type system is strong enough to imply termination, this manifests directly as every term being given a value. If non-terminating terms are possible, it will be necessary to introduce, as part of the semantics, special cases to deal with that,\cite[\S 5.2]{MOSSES1990575} otherwise the definition will simply not be consistent.

@@

There are a few variations on how the traces work. In some versions (as is the case in Section \ref{operational semantics}, and \cite{borgstrom2016lambda}), each run of the program is assigned a trace of finite length, containing exactly those samples actually used in that run. If the set of results for an individual sampling operation is $S$ (in this case it's the set of 2 elements, but in a language with continuous distributions, it could be $\mathbb R$, or an interval, or something else), the set of traces is $\bigcup_{n \in \mathbb N} S^n$, and the measure defined on the set of traces is not actually a probability measure, as each set $S^n$ has measure 1. This still acts somewhat like a probability measure though, as traces of different lengths are mutually exclusive. If a trace $a$ is a prefix of a trace $b$ (with $b = a + c$), then it is not possible that $(M, a, 1) \to^* (V, [], s)$ and also $(M, b, 1) \to^* (V', [], s')$, because the evaluation with any given trace is deterministic, and $(M, b, 1) \to^* (V, c, s)$, at which point it could not proceed any further, having already reached a value. For any term, the set of traces for which it terminates therefore has measure at most 1.

Alternatively, the trace could contain infinitely many values, with any finite execution of a program leaving some of the samples unused. This is in some ways a little simpler, because the set of traces can simply have a probability distribution, and the trace does not need to depend in any way on the program (by having its length matched). An additional variant of this style of trace, used for example in \cite{culpepper17}, is to split the trace in three parts, all infinitely long, for the application case of the big-step semantics, rather than having the each step (evaluating the function, evaluating the argument, then evaluating the result of the $\beta$ reduction) use however many samples it needs from the start of the trace in turn.

@@

The distinction between distribution and trace-based semantics is discussed in \cite{borgstrom2016lambda}, which gives both for the same language, and proves their equivalence (in the sense that the distributions derived from the trace-based semantics are the same as the distributions that the distribution-based semantics produces directly).

@@

The second extension is antitone ranking functions. This is closely related to the progress condition given in \cite{mciver2017new}, just adapted to the new functional context. A function $\epsilon : \mathbb R_{\geq 0} \to \mathbb R_{> 0}$, which is required to be monotoically decreasing, is added to the definition, and rather than the ranking function having to decrease by 1 at every $\tY$-reduction step, it only has to decrease by $\epsilon$ of the current value of the ranking function. This allows the termination of programs which terminate much more slowly (such as the unbiased random walk) to be proven, whereas the basic result was only applicable to terms which terminated with a finite expected number of $\tY$-reduction steps.

@@

\subsection{Completeness of Antitone Rankability}
A term is rankable iff every term reachable from it terminates with finitely many $\tY$-reduction steps in expectation. This essentially puts a limit on how slowly a term can terminate while still being rankable, so any term which is not $\tY$-positive almost surely terminating cannot be proven AST by assigning it a ranking function. With antitone ranking functions, there is no analogous restriction: there are arbitrarily slowly terminating terms (i.e. for any probability distribution on $\mathbb N$, there is a term whose distribution of number of $\tY$-reduction steps before termination is the given distribution) which are antitone rankable. It therefore seems rather likely that any term such that every term reachable from is is AST is antitone rankable. I intend to try to prove this result, even though it does not relate very closely to my overall thesis topic, because it would make my existing work more complete.

\bibliography{references}
\end{document}
